|

### 2023-11-10 src/docs_paper/paper.Rmd
small typo
Remaining paper and appendix changes

---
title: |
  | Beyond borders: evaluating the suitability of spatial adjacency for small-area estimation
classoption: ba
author:
  - firstname: Adam 
    lastname: Howes
    email: ath19@ic.ac.uk
    affiliationref: A,B
    footnoterefs: 
      - t1
  - firstname: Jeffrey W.
    lastname: Eaton
    affiliationref: C,B
  - firstname: Seth R.
    lastname: Flaxman
    affiliationref: D
affiliations:
  - ref: A
    name: Imperial College London 
    address: Department of Mathematics
  - ref: B
    name: Imperial College London
    address: MRC Centre for Global Infectious Disease Analysis
  - ref: C
    name: Harvard University
    address: Center for Communicable Disease Dynamics, Harvard T.H. Chan School of Public Health
  - ref: D
    name: University of Oxford
    address: Department of Computer Science 
footnotes:
  - ref: t1
    text: "Corresponding Author"
supplements:
  - title: Appendix
    description: All further materials for this study.
acknowledgements: |
  | AH thanks Robert Ashton for help with \textsf{R} package development, and Harrison Zhu, Theo Rashid, Tim Wolock, Samir Bhatt, Tim Lucas, Elizaveta Semenova, Marta Blangiardo and Oliver Ratmann for helpful conversations. Yidan Xu undertook the Mary Lister McCammon Summer Research Fellowship 2019 on related work.
abstract: |
  |  Spatial correlation between small-areas is typically accounted for using spatial random effects. However, commonly used models for spatial random effects based upon adjacency relations, like the Besag model, make unrealistic spatial assumptions for geometries which are not a grid. Furthermore, they do not take into account areal data is commonly generated by aggregating continuous geostatistical data. Defining spatial random effects which instead correspond to aggregated Gaussian processes gives a more intuitively convincing correlation structure, and allows tools from kernel methods to be used in models for areal data.
  |
  |  In a simulation study, we used a proper scoring rule to evaluate the performance of different spatial random effect specifications. The simulations were performed under model misspecification and across a range of vignette and realistic geometries with different regularities. We also compared model performance in estimating district-level HIV prevalence from PHIA household survey data across four countries in sub-Saharan Africa, using cross-validation and spatial cross-validation.
  |
  |  The results of this analysis are forthcoming, and will be described here.
date: "`r Sys.Date()`"
bibliography: citations.bib
#linenumbers: true
#numbersections: true
output: 
  bookdown::pdf_book: 
    base_format: rticles::isba_article
    md_extensions: -autolink_bare_uris #keep these off or \printead{} will fail on email addresses
    includes:
      in_header: preamble.tex
---

@@

Often, as well as spatial structure, there exists IID over-dispersion in the residuals and it is inappropriate to use purely spatially structured random effects in the model.
The Besag-York-Mollié (BYM) model of @besag1991bayesian, accounts for this in a natural way by decomposing the spatial random effect $\bphi = \bv + \bu$ into a sum of an unstructured IID component $\bv$ and a spatially structured Besag component $\bu$, each of which with their own respective precision parameters $\tau_v$ and $\tau_u$.
The resulting distribution is
\begin{equation}
    \bphi \sim \N(0, \tau_v^{-1} \mathbf{I} + \tau_u^{-1} \mathbf{R}^{-}) \label{eq:bym}.
\end{equation}
Including both $\bv$ and $\bu$ is intended to enable the model to learn the relative extent of the unstructured and structured components via $\tau_v$ and $\tau_u$.
However, in this specification, scaling of the Besag precision matrix $\mathbf{Q}$ is not taken into account despite this issue being particularly pertinent when dealing with multiple sources of noise.
In particular, placing a joint prior $(\tau_u, \tau_v) \sim p(\tau_u, \tau_v)$ which doesn't privilege either component is more easily accomplished if $\mathbf{Q}$ and $\mathbf{I}$ have the same scale.
Additionally, supposing we have a prior belief that the over-dispersion is primarily IID and $\bv$ accounts for the majority of the dispersion, then it is not immediately obvious how to represent this belief using $p(\tau_u, \tau_v)$, without inadvertently altering the prior about the overall variation.
This highlights identifiability issues of the parameters $(\tau_u, \tau_v)$ resulting from them not being orthogonal.
Building on the models of @leroux2000estimation and @dean2001detecting which tackle this the identifiability problem, but do not scale the spatially structured noise, @simpson2017penalising propose a reparameterisation $(\tau_v, \tau_u) \mapsto (\tau_\phi, \pi)$ of the BYM model known as the BYM2 model and given by
\begin{align}
\bphi = \frac{1}{\tau_\phi} \left( \sqrt{1- \pi} \, \bv + \sqrt{\pi} \, \bu^\star \right), \label{eq:bym2}
\end{align}
where $\tau_\phi$ is the marginal precision of $\bphi$, $\pi \in [0, 1]$ gives the proportion of the marginal variance explained by each component, and $\bu^\star$ is a scaled version of $u$ with precision matrix given by the scaled structure matrix $\mathbf{R}^\star$ (see Appendix).
When $\pi = 0$ the random effects are IID, and when $\pi = 1$ the random effects follow the Besag model.
To borrow an analogy [@rue2020comment] the parameterisation $(\tau_v, \tau_u)$ is like having one hot water and one cold water tap, whereas the parameterisation $(\tau_\phi, \pi)$ is like a mixer tap where the amount of water and its temperature can be adjusted separately.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{depends/simulation-geometries.pdf}
\caption{The seven simulation geometries that we considered.}
\label{fig:51}
\end{figure}

@@

\begin{table}
\small
\begin{tabularx}{\textwidth}{llX}
\toprule
& Inferential model & Additional details \\
 \midrule
\hypertarget{I1}{\texttt{I1}} & Constant &
No spatial random effects \label{row:I1} \\
\hypertarget{I2}{\texttt{I2}} & IID &
$\bphi \sim \mathcal{N}(0, \tau_\phi^{-1} \mathbf{I}_n)$\\
\hypertarget{I3}{\texttt{I3}} & Besag &
$\bphi \sim \mathcal{N}(0, (\tau_\phi \mathbf{R}^\star)^{-})$ following additional recommendations of \citet{freni2018note}  \\
\hypertarget{I4}{\texttt{I4}} & BYM2 &
$\bphi = \tau_\phi^{-1} \left( \sqrt{1- \pi} \, \bv + \sqrt{\pi} \, \bu^\star \right)$ with a $\text{Beta}(0.5, 0.5)$ prior on $\pi \in [0, 1]$ \\
\hypertarget{I5}{\texttt{I5}} & FCK &
$\bphi \sim \mathcal{N}(0, \tau_\phi^{-1} \mathbf{K})$ with $K_{ij} = k(c_i, c_j)$ where the length-scale $l$ is fixed \\
\hypertarget{I6}{\texttt{I6}} & CK & As \hypertarget{I5}{\texttt{I5}}, with length-scale prior $l \sim \mathrm{Inv{\text-}Gamma}(a, b)$ \\
\hypertarget{I7}{\texttt{I7}} & FIK & $\bphi \sim \mathcal{N}(0, \tau_\phi^{-1} \mathbf{K})$ with $K_{ij} = \frac{1}{10^2} \sum_{l = 1}^{10} \sum_{m = 1}^{10} k \left( s^{(i)}_l, s^{(j)}_m \right)$ \newline with hexagonal integration point spacing and fixed length-scale $l$ \\
\hypertarget{I8}{\texttt{I8}} & IK & As \hypertarget{I7}{\texttt{I7}}, with length-scale prior $l \sim \mathrm{Inv{\text-}Gamma}(a, b)$ \\
\bottomrule
\end{tabularx}
\caption{All inferential models are implemented as a part of the \textsf{R} package \texttt{arealutils} \citep{arealutils}. The AGHQ algorithm was used for approximate Bayesian inference of all inferential models, via the \texttt{aghq} \textsf{R} package \citep{stringer2021implementing}}
\label{tab:inferentialmodels}
\small
\end{table}

@@

We used adaptive Gauss-Hermite quadrature [AGHQ; @stringer2022fast] with $k = 3$ quadrature points per hyperparameter dimension to perform approximate Bayesian inference.
Each model was implemented using a Template Model Builder \textsc{C++} template for the log-posterior via the `TMB` package [@kristensen2016tmb]
We were then able to conduct inference using both AGHQ via the `aghq` package [@stringer2021implementing] and the No-U-Turn Sampling (NUTS) Hamiltonian Monte Carlo (HMC) algorithm using Stan [@carpenter2017stan] via the `tmbstan` package [@monnahan2018no].
In Appendix we demonstrate that the inference results from AGHQ are comparable to those obtained running NUTS over a longer time.
We also compare to empirical Bayes, and where possible integrated nested Laplace approximation via `R-INLA` [@martins2013bayesian].

@@

We assessed each fitted inferential model according to its ability to recover the true underlying value of the probabilities $\rho_i$ at each area in the study region, as well as the calibration of the model's estimates.
<!-- We describe our approach to model assessment in the following sections. -->
<!-- Analogous approaches may be used to check recovery of the intercept parameter $\beta_0$, precision parameter $\tau_\phi$, and, when relevant, the length-scale $l$. -->

@@

The continuous ranked probability score [CRPS; @matheson1976scoring] generalises the Brier score [@brier1950verification] to distributional forecasts.
Let $\rho_i$ have posterior marginal $f(\rho_i) = p(\rho_i \, | \, y)$ and $\omega_i$ be the true value.
Writing $F$ as the cumulative distribution function corresponding to $f$ then
\begin{equation}
\text{CRPS}(f, \omega_i) = \int_{-\infty}^{\infty} (F(\rho_i) - \mathbbm{1} \{\rho_i \geq \omega_i \} )^2 \text{d}\rho_i = \int_{0}^{1} (F(\rho_i) - \mathbbm{1} \{\rho_i \geq \omega_i \} )^2 \text{d}\rho_i, \label{eq:crps}
\end{equation}
where $\mathbbm{1}$ denotes the indicator function and the second equality follows from $0 \leq \rho_i \leq 1$.
$\text{CRPS}(f, \omega_i)$ if and only if $f(\rho_i) = \delta_{\omega_i}$.
The CRPS is a strictly proper scoring rule^[As is the logarithmic score [@good1952rational] given by $\text{LogS}(f, \omega_i) = \log f(\omega_i)$, though in line with the findings of @kruger2020predictive, we found it difficult to compute satisfactorily using kernel density estimation.] [@gneiting2007strictly] and can be evaluated directly using samples $\{\rho_i^s\}_{s = 1}^S$ as
\begin{equation}
\text{CRPS}(f, \omega_i) \approx \frac{1}{S} \sum_{s = 1}^S | \rho_i^s - \omega_i | - \frac{1}{2S^2} \sum_{s = 1}^S \sum_{l = 1}^S | \rho_i^s - \rho_i^l |. \label{eq:crpsapprox}
\end{equation}

@@

See Appendix.

@@

>>>>

---
title: |
  | Beyond borders: evaluating the suitability of spatial adjacency for small-area estimation
runtitle: "Evaluating the suitability of spatial adjacency for small-area estimation"
classoption: ba
author:
  - firstname: Adam 
    lastname: Howes
    email: ath19@ic.ac.uk
    affiliationref: A,B
    footnoterefs: 
      - t1
  - firstname: Jeffrey W.
    lastname: Eaton
    affiliationref: C,B
  - firstname: Seth R.
    lastname: Flaxman
    affiliationref: D
affiliations:
  - ref: A
    name: Imperial College London 
    address: Department of Mathematics
  - ref: B
    name: Imperial College London
    address: MRC Centre for Global Infectious Disease Analysis
  - ref: C
    name: Harvard University
    address: Center for Communicable Disease Dynamics, Harvard T.H. Chan School of Public Health
  - ref: D
    name: University of Oxford
    address: Department of Computer Science 
footnotes:
  - ref: t1
    text: "Corresponding Author"
supplements:
  - title: Appendix
    description: All further materials for this study.
acknowledgements: |
  | AH thanks Robert Ashton for help with \textsf{R} package development, and Harrison Zhu, Theo Rashid, Tim Wolock, Samir Bhatt, Tim Lucas, Elizaveta Semenova, Marta Blangiardo and Oliver Ratmann for helpful conversations. Yidan Xu undertook the Mary Lister McCammon Summer Research Fellowship 2019 on related work.
abstract: |
  |  Spatial correlation between small-areas is typically accounted for using spatial random effects. However, commonly used models for spatial random effects based upon adjacency relations, like the Besag model, make unrealistic spatial assumptions for geometries which are not a grid. Furthermore, they do not take into account areal data is commonly generated by aggregating continuous geostatistical data. Defining spatial random effects which instead correspond to aggregated Gaussian processes gives a more intuitively convincing correlation structure, and allows tools from kernel methods to be used in models for areal data.
  |
  |  In a simulation study, we used a proper scoring rule to evaluate the performance of different spatial random effect specifications. The simulations were performed under model misspecification and across a range of vignette and realistic geometries with different regularities. We also compared model performance in estimating district-level HIV prevalence from PHIA household survey data across four countries in sub-Saharan Africa, using cross-validation and spatial cross-validation.
  |
  |  The results of this analysis are forthcoming, and will be described here.
date: "`r Sys.Date()`"
bibliography: citations.bib
linenumbers: true
#numbersections: true
output: 
  bookdown::pdf_book: 
    base_format: rticles::isba_article
    md_extensions: -autolink_bare_uris
    includes:
      in_header: preamble.tex
---

@@

Often, as well as spatial structure, there exists IID over-dispersion in the residuals and it is inappropriate to use purely spatially structured random effects in the model.
The Besag-York-Mollié (BYM) model of @besag1991bayesian, accounts for this in a natural way by decomposing the spatial random effect $\bphi = \bv + \bu$ into a sum of an unstructured IID component $\bv$ and a spatially structured Besag component $\bu$, each of which with their own respective precision parameters $\tau_v$ and $\tau_u$.
The resulting distribution is
\begin{equation}
    \bphi \sim \N(0, \tau_v^{-1} \mathbf{I} + \tau_u^{-1} \mathbf{R}^{-}) \label{eq:bym}.
\end{equation}
Including both $\bv$ and $\bu$ is intended to enable the model to learn the relative extent of the unstructured and structured components via $\tau_v$ and $\tau_u$.
However, in this specification, scaling of the Besag precision matrix $\mathbf{Q}$ is not taken into account despite this issue being particularly pertinent when dealing with multiple sources of noise.
In particular, placing a joint prior $(\tau_u, \tau_v) \sim p(\tau_u, \tau_v)$ which doesn't privilege either component is more easily accomplished if $\mathbf{Q}$ and $\mathbf{I}$ have the same scale.
Additionally, supposing we have a prior belief that the over-dispersion is primarily IID and $\bv$ accounts for the majority of the dispersion, then it is not immediately obvious how to represent this belief using $p(\tau_u, \tau_v)$, without inadvertently altering the prior about the overall variation.
This highlights identifiability issues of the parameters $(\tau_u, \tau_v)$ resulting from them not being orthogonal.
Building on the models of @leroux2000estimation and @dean2001detecting which tackle this the identifiability problem, but do not scale the spatially structured noise, @simpson2017penalising propose a reparameterisation $(\tau_v, \tau_u) \mapsto (\tau_\phi, \pi)$ of the BYM model known as the BYM2 model and given by
\begin{align}
\bphi = \frac{1}{\tau_\phi} \left( \sqrt{1- \pi} \, \bv + \sqrt{\pi} \, \bu^\star \right), \label{eq:bym2}
\end{align}
where $\tau_\phi$ is the marginal precision of $\bphi$, $\pi \in [0, 1]$ gives the proportion of the marginal variance explained by each component, and $\bu^\star$ is a scaled version of $u$ with precision matrix given by the scaled structure matrix $\mathbf{R}^\star$ (see Appendix S1).
When $\pi = 0$ the random effects are IID, and when $\pi = 1$ the random effects follow the Besag model.
To borrow an analogy [@rue2020comment] the parameterisation $(\tau_v, \tau_u)$ is like having one hot water and one cold water tap, whereas the parameterisation $(\tau_\phi, \pi)$ is like a mixer tap where the amount of water and its temperature can be adjusted separately.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{depends/simulation-geometries.pdf}
\caption{The seven simulation geometries that we considered. The top row contains the four vignette geometries from \ref{fig:tikz1}, and the bottom row contains three more realistic geometries.}
\label{fig:51}
\end{figure}

@@

\begin{table}
\small
\begin{tabularx}{\textwidth}{llX}
\toprule
& Inferential model & Additional details \\
 \midrule
\hypertarget{I1}{\texttt{I1}} & Constant &
No spatial random effects \label{row:I1} \\
\hypertarget{I2}{\texttt{I2}} & IID &
$\bphi \sim \mathcal{N}(0, \tau_\phi^{-1} \mathbf{I}_n)$\\
\hypertarget{I3}{\texttt{I3}} & Besag &
$\bphi \sim \mathcal{N}(0, (\tau_\phi \mathbf{R}^\star)^{-})$ following additional recommendations of \citet{freni2018note} (Appendix S1)  \\
\hypertarget{I4}{\texttt{I4}} & BYM2 &
$\bphi = \tau_\phi^{-1} \left( \sqrt{1- \pi} \, \bv + \sqrt{\pi} \, \bu^\star \right)$ with a $\text{Beta}(0.5, 0.5)$ prior on $\pi \in [0, 1]$ \\
\hypertarget{I5}{\texttt{I5}} & FCK &
$\bphi \sim \mathcal{N}(0, \tau_\phi^{-1} \mathbf{K})$ with $K_{ij} = k(c_i, c_j)$ where the length-scale $l$ is fixed \\
\hypertarget{I6}{\texttt{I6}} & CK & As \hypertarget{I5}{\texttt{I5}}, with length-scale prior $l \sim \mathrm{Inv{\text-}Gamma}(a, b)$ \\
\hypertarget{I7}{\texttt{I7}} & FIK & $\bphi \sim \mathcal{N}(0, \tau_\phi^{-1} \mathbf{K})$ with $K_{ij} = \frac{1}{10^2} \sum_{l = 1}^{10} \sum_{m = 1}^{10} k \left( s^{(i)}_l, s^{(j)}_m \right)$ \newline with hexagonal integration point spacing and fixed length-scale $l$ \\
\hypertarget{I8}{\texttt{I8}} & IK & As \hypertarget{I7}{\texttt{I7}}, with length-scale prior $l \sim \mathrm{Inv{\text-}Gamma}(a, b)$ \\
\bottomrule
\end{tabularx}
\caption{All inferential models are implemented as a part of the \textsf{R} package \texttt{arealutils} \citep{arealutils}. We used adaptive Gauss-Hermite quadrature for approximate Bayesian inference, implemented via the \texttt{aghq} \textsf{R} package \citep{stringer2021implementing}}
\label{tab:inferentialmodels}
\small
\end{table}

@@

We used adaptive Gauss-Hermite quadrature [AGHQ; @stringer2022fast] with $k = 3$ quadrature points per hyperparameter dimension to perform approximate Bayesian inference.
Each model was implemented using a Template Model Builder \textsc{C++} template for the log-posterior via the `TMB` package [@kristensen2016tmb].
We were then able to conduct inference using both AGHQ via the `aghq` package [@stringer2021implementing] and the No-U-Turn Sampling (NUTS) Hamiltonian Monte Carlo (HMC) algorithm using Stan [@carpenter2017stan] via the `tmbstan` package [@monnahan2018no].
In Appendix S2 we demonstrate that the inference results from AGHQ are comparable to those obtained running NUTS to convergence.
<!-- We also compare to empirical Bayes, and where possible integrated nested Laplace approximation via `R-INLA` [@martins2013bayesian]. -->

@@

We assessed each fitted inferential model according to its ability to recover the true underlying value of $\rho_i$ at each area in the study region, as well as the calibration of the model's estimates.
<!-- We describe our approach to model assessment in the following sections. -->
<!-- Analogous approaches may be used to check recovery of the intercept parameter $\beta_0$, precision parameter $\tau_\phi$, and, when relevant, the length-scale $l$. -->

@@

The continuous ranked probability score [CRPS; @matheson1976scoring] generalises the Brier score [@brier1950verification] to distributional forecasts.
Let $\rho_i$ have posterior marginal $f(\rho_i) = p(\rho_i \, | \, y)$ and $\omega_i$ be the true value.
Writing $F$ as the cumulative distribution function corresponding to $f$ then
\begin{equation}
\text{CRPS}(f, \omega_i) = \int_{-\infty}^{\infty} (F(\rho_i) - \mathbbm{1} \{\rho_i \geq \omega_i \} )^2 \text{d}\rho_i = \int_{0}^{1} (F(\rho_i) - \mathbbm{1} \{\rho_i \geq \omega_i \} )^2 \text{d}\rho_i, \label{eq:crps}
\end{equation}
where $\mathbbm{1}$ denotes the indicator function and the second equality follows from $0 \leq \rho_i \leq 1$.
$\text{CRPS}(f, \omega_i)$ if and only if $f(\rho_i) = \delta_{\omega_i}$.
The CRPS is a strictly proper scoring rule and can be evaluated directly using samples^[The logarithmic score [@good1952rational] given by $\text{LogS}(f, \omega_i) = \log f(\omega_i)$ is another strictly proper scoring rule, however we found it difficult to compute satisfactorily using kernel density estimation, in line with the findings of @kruger2020predictive.] [@gneiting2007strictly] $\{\rho_i^s\}_{s = 1}^S$ as
\begin{equation}
\text{CRPS}(f, \omega_i) \approx \frac{1}{S} \sum_{s = 1}^S | \rho_i^s - \omega_i | - \frac{1}{2S^2} \sum_{s = 1}^S \sum_{l = 1}^S | \rho_i^s - \rho_i^l |. \label{eq:crpsapprox}
\end{equation}

@@

See Appendix S3.

@@

See Appendix S4.
